Algorithm explanation

### 2013-08-22 article/algo.tex
Explanation of Horn graph

We propose a new algorithm for solving Horn clauses to generate a
solution which is as simple as possible.  This enables us to obtain
appropriate predicates that are suitable for predicate abstraction in
the program verification techniques.

The construction of the algorithm roughly consists of three parts.
\begin{itemize}
\item \textbf{Horn Graph Building} converts input Horn clauses into a
  \emph{Horn graph} that express the equivalent problem. The process
  also generates a predicate template for every predicate variable
  that the solution will become.  The Horn graph is used throughout
  the algorithm to represent the problem under certain solution
  template.
\item \textbf{Iterative Constraint Generation} is a process to build a
  constraint to compute a solution.  It starts from generating the
  weakest constraint in the sense that the resulting solution would not
  be simple yet.  The process iteratively strengthen a constraint
  until the strongest one while preserving the constraint
  satisfiability. The strongest constraint gives an instanciation of
  templates.
\item \textbf{Graph Restructuring} If the constraint become
  unsatisfiable during the iteration, the process restructure the
  input Horn graph to enlarge the solution template.  This is done by
  considering the unsatisfiable core of the constraint.
\end{itemize}
Our algorithm works in a constructive manner.

Each predicate variable vertex in an input graph has a unique
predicate variable label.  All predicate variables on them are
assigned simple templates.  Those predicate templates are expressed
with template parameters.  Then, according to the graph structure, the
algorithm builds a linear constraint that parameters must satisfy.
For constraint building, the algorithm traverses the input graph in
the topological order on $V_\Rightarrow$, and concatenate the
constraints from predecessors to build the problem constraints.  Each
process to build a constraint on $\mathcal{HC}_v$ for all vertices $v
\in V_\Rightarrow$ is equivalent to generate a linear constraint in
interpolation.

@@

\paragraph {DAG Horn graph solving}
Second, we explain a case that some predicate variables appear
multiple times on left-hand side in different Horn clauses.

@@

\paragraph {Disjunctive Horn graph solving}
Finally, we explain a case that an input set of Horn clauses contain
disjunctions on left-hand side.

@@

@@

Our Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed acyclic graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two disjoint sets of vertices; Horn term
  vertices $V_T$ and arrow vertices $V_\Rightarrow$. Each vertex $u
  \in V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, \ldots, x_{\mathrm{arity}(P)})$ or a linear
  inequality $a_1 x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}
There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, written $\edgel{u}{\theta}{v}$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, written $\edge{u}{v}$, where the mapping is
  empty.
\end{itemize}

The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses $\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$.

@@

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $\textsc{FV}(\mathcal{HC}_v)$.  Without loss
of generality, we assume the existence of the root vertex $v_\bot \in
V_T$ with linear expression labels $1 \leq 0$, meaning $\bot$. We also
assume that other vertices with linear expression labels have no
incoming edges to them.

@@

We first propose an algorithm to compute a simple solution for the
given $G$ if one exists.  Later we add modifications to the algorithm
for the case that a simple solution does not exist.  Algorithm~\ref{alg:solveHorn} shows the pseudo-code of our algorithm.

@@

>>>>

We propose a new algorithm for solving recursion-less Horn clauses to
generate a solution which is as simple as possible.  This enables us
to obtain appropriate predicates that are suitable for predicate
abstraction in the program verification techniques.

The construction of the algorithm roughly consists of two parts.
\begin{itemize}
\item \textbf{Iterative Constraint Generation} is a process to build a
  constraint to compute a solution.  It starts from generating the
  weakest constraint in the sense that the resulting solution would
  not be simple yet.  The process iteratively strengthen a constraint
  until the strongest one while preserving the constraint
  satisfiability.  The strongest constraint gives a simple solution.
\item \textbf{Problem Restructuring} If the constraint become
  unsatisfiable during the iteration, the process restructure the
  problem equivalent to obtain a less simpler solution.  This process
  considers the unsatisfiable core of the constraint.
\end{itemize}

All predicate variables in an input problem are assigned simple
templates.  Those predicate templates are parameterized.  In the
iterative constraint generation process, algorithm builds a linear
constraint that parameters in templates must satisfy.  This is done in
a constructive manner, and constraints are build in the order of
implication dependency in Horn clauses.  That is, when an input
problem has a Horn clause thath has a predicate variable $P$ on its
left-hand and $Q$ on its right-hand, a constraint to describe
parameters in $P$'s template are first build before those in $Q$'s
template.  When all parameters in the whole problem are described, the
built constraint are evaluated to obtain a model.  Its instanciation
becomes an appropriate solution for the input problem.

The novelty of our algorithm lies in the incremental constraint
refinement.  The algorithm try from the weakest constraint that an
input problem allows that does not give a simple solution, and
strengthen the constraint until it reaches to the simplest solution
that the problem has.  When a constraint becomes unsatisfiable, the
algorithm analyzes the unsatisfiable core and transform the problem to
obtain less simple solution.

This construction of the algorithm enables us to compute a relatively
simpler solution compared to existing algorithms in a reasonable time.

@@

\paragraph {DAG problem solving}
Second, we explain a case that some predicate variables appear
multiple times on left-hand side in different Horn clauses.

@@

\paragraph {Disjunctive Horn clause solving}
Finally, we explain a case that an input set of Horn clauses contain
disjunctions on left-hand side.

@@

The process also generates a predicate template for every predicate
variable that the solution will become.  under certain solution
template.

@@

For ease of discussion, our algorithm handles an input set of Horn
clauses by a \emph{Horn graph} that express the equivalent problem.
The graph is then used throughout the algorithm to represent the
problem.  A Horn graph is a labelled directed acyclic graph
$G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two disjoint sets of vertices; Horn term
  vertices $V_T$ and arrow vertices $V_\Rightarrow$.  Each vertex $u
  \in V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, \ldots, x_{\mathrm{arity}(P)})$ or a linear
  inequality $a_1 x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}
There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, written $\edgel{u}{\theta}{v}$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, written $\edge{u}{v}$, where the mapping is
  empty.
\end{itemize}
The meaning $\llbracket G \rrbracket $ of a Horn graph $G$ is a set of
Horn clauses $\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$.

@@

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $\textsc{FV}(\mathcal{HC}_v)$.

Without loss of generality, we assume the existence of the root vertex
$v_\bot \in V_T$ with linear expression labels $1 \leq 0$, meaning
$\bot$.  This is because if we have a Horn clause $P \implies e$ which
$e$ is a linear expression, we can convert it to $P \wedge \neg e
\implies 1 \leq 0$.  We also assume that other vertices with linear
expression labels have no incoming edges to them.

@@

We first propose an algorithm to compute a simple solution for the
given Horn graph $G$ if one exists.  Later we add modifications to the
algorithm for the case that a simple solution does not exist for $G$.
Algorithm~\ref{alg:solveHorn} shows the pseudo-code of our algorithm.

@@

traverses the input graph in the topological order on
$V_\Rightarrow$, and concatenate the constraints from predecessors to
build the problem constraints.  Each process to build a constraint on
$\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$ is equivalent
to generate a linear constraint in interpolation.
